A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-26 jour/jour.tex
tuning

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following hold for all $\ec{x}$ and $\ec{y}$ in the domain of $\ValC$:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.
\item[\rm (2)]
$\ValC \ec{x} = \expand{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}

@@

\begin{examplex}
\label{ex:assign-complete}
Consider $\Val$ from Example~\ref{ex:assign}.
While extending $\Val$ to its corresponding completion $\ValC$,
we may, for instance, assign $\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ to $\Varec{z}$,
and subsequently assign $\MU {\Varec{y}}.\; \const{D}( {\Varec{y}} )$ to $\Varec{y}$.
We obtain the following mapping $\ValC$ from equivalence classes to $\MU$-terms:
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\ValC \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\ValC \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\ValC \ec{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\ValC \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\ValC \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\ValC \ec{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
Notice that $\ValC$ satisfies the properties from Lemma~\ref{lem:interpretation-completion} for a normal interpretation $\J$:
all terms in the range of $\ValC$ are closed and normal, 
$\ValC \ec{s} = \expand{\ValC \ec{y}}{ t }$ for all subterms $t$ where $\muvar(t) = {\Varec{s}}$ for all $s$,
and all terms in the range of $\ValC$ are pairwise alpha-disequivalent.
We may construct an interpretation $\J$ where variables are interpreted as the value
that is alpha-equivalent to one from $\interp{\J}{\delta}$, and is uniquely associated to its equivalence class by $\ValC$.
%For instance, up to renaming,
%$\interp{\J}{u}$ is $\const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) )$,
%$\interp{\J}{v}$ and $\interp{\J}{s}$ are $\const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) )$, and so on.
\xend
\end{examplex}

@@

\begin{examplex}
\label{ex:model}
Continuing Example~\ref{ex:assign-complete}, let $\J$ be a normal interpretation, where
up to renaming of $\MU$-bound variables, we have:
\[\begin{array}{r@{}c@{}r@{}c@{}l@{\qquad}r@{}c@{}l@{\qquad}r@{}c@{}l}
& &\interp{\J}{u} & {}\,=\,{} &  \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{w} & = &  \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\interp{\J}{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\interp{\J}{v} & = & \interp{\J}{s} & = & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{x} & {}\,=\,{} & \const{C}( \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\interp{\J}{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
It can be shown $\J$ satisfies $\Ec =
\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
v \teq s,\;
z \not\teq v
\}$.
\xend
\end{examplex}

@@

Assume $\Ec$ is initially the set $M_i$ described above. With each equality $t \teq u$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq u$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is assigned an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
In the case that multiple rules could derive the same conclusion,
the implementation simply keeps the first explanation it encountered.
The rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. The implementation does not 
use any techniques to minimize the size of this set, which in some cases may be non-minimal.
For \rn{Split}, we add the exhaustiveness clause %of the form
$\DISC{1} \mathrel\lor \cdots \mathrel\lor \DISC{m}$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

@@

The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
Discriminators are part of the signature and not abbreviations.
This requires extending the decision procedure with several
rules, % \cite{barrett-et-al-2007},
which apply uniformly to datatypes and codatatypes.
This approach often leads to better performance because it introduces terms
less eagerly to $\tEc$.
Selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, the solver directly adds
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality.
To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is smaller than $u$ with respect to some term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
Finally, finite ordinary types are not converted to finite datatypes.
We instead rely on standard extensions of Nelson-Oppen theory combination~\cite{tinelli2005combining}, which
impose requirements regarding terms having finite sorts are shared
between multiple theories.  In particular, given an ordinary type $\tau$ of finite cardinality $n$
belonging to theory $T_i$, the theory solver for $T_i$ must 
communicate enough equalities over shared terms to
ensure all other theory solvers, including the one for $\thD$, 
interpret $\tau$ as a set having cardinality at most $n$.

>>>>

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following
properties hold for all $\ec{x}$ and $\ec{y}$ in the domain of $\ValC$:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.
\item[\rm (2)]
$\ValC \ec{x} = \expand{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}

@@

\begin{examplex}
\label{ex:assign-complete}
Consider $\Val$ from Example~\ref{ex:assign}.
While extending $\Val$ to its corresponding completion $\ValC$,
we may, for instance, assign $\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ to $\Varec{z}$,
and subsequently assign $\MU {\Varec{y}}.\; \const{D}( {\Varec{y}} )$ to $\Varec{y}$.
We obtain the following mapping $\ValC$ from equivalence classes to $\MU$-terms:
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\ValC \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\ValC \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\ValC \ec{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\ValC \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\ValC \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\ValC \ec{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
Notice that $\ValC$ satisfies the properties from Lemma~\ref{lem:interpretation-completion}
for a normal interpretation $\J$:
All terms in the range of $\ValC$ are closed and normal,
$\ValC \ec{s} = \expand{\ValC \ec{y}}{ t }$ for all subterms $t$ where $\muvar(t) = {\Varec{s}}$ for all $s$,
and all terms in the range of $\ValC$ are pairwise $\alpha$-disequivalent.
We may construct an interpretation $\J$ where variables are interpreted as the value
that is $\alpha$-equivalent to one from $\interp{\J}{\delta}$ and that is uniquely associated
with its equivalence class by $\ValC$.
%For instance, up to renaming,
%$\interp{\J}{u}$ is $\const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) )$,
%$\interp{\J}{v}$ and $\interp{\J}{s}$ are $\const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) )$, and so on.
\xend
\end{examplex}

@@

\begin{examplex}
\label{ex:model}
Continuing Example~\ref{ex:assign-complete}, let $\J$ be a normal interpretation, where
up to renaming of $\MU$-bound variables, we have
\[\begin{array}{r@{}c@{}r@{}c@{}l@{\qquad}r@{}c@{}l@{\qquad}r@{}c@{}l}
& &\interp{\J}{u} & {}\,=\,{} &  \const{C}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{w} & = &  \const{E}( \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) ) &
\interp{\J}{y} & {}\,=\,{} & \MU {\Varec{y}}.\; \const{D}( {\Varec{y}} ) \\
\interp{\J}{v} & = & \interp{\J}{s} & = & \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) &
\interp{\J}{x} & {}\,=\,{} & \const{C}( \const{D}( \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} ) ) ) &
\interp{\J}{z} & = & \MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )
\end{array}\]
It can be shown that $\J$ satisfies $\Ec =
\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
v \teq s,\;
z \not\teq v
\}$.
\xend
\end{examplex}

@@

Assume $\Ec$ is initially the set $M_i$ described above. With each equality $t \teq u$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq u$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is assigned an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
If multiple rules can derive the same conclusion,
the implementation simply keeps the first explanation it encounters.

The rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. The implementation does not
use any techniques to minimize the size of this set, which in some cases may be non-minimal.
For \rn{Split}, we add the exhaustiveness clause %of the form
$\DISC{1} \mathrel\lor \cdots \mathrel\lor \DISC{m}$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

@@

The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
Discriminators are part of the signature and not abbreviations.
This requires extending the decision procedure with several
rules, % \cite{barrett-et-al-2007},
which apply uniformly to datatypes and codatatypes.
This approach often leads to better performance because it introduces terms
less eagerly to $\tEc$.
Selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, the solver directly adds
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality.
To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is smaller than $u$ with respect to some term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
Finally, finite ordinary types are not converted to finite datatypes.
We instead rely on standard extensions of the Nelson--Oppen method~\cite{tinelli2005combining}, which
impose requirements regarding terms having finite sorts are shared
between multiple theories.  In particular, given an ordinary type $\tau$ of finite cardinality $n$
belonging to theory $T_i$, the theory solver for $T_i$ must
communicate enough equalities over shared terms to
ensure that all other theory solvers, including the one for $\thD$,
interpret $\tau$ as a set having cardinality at most $n$.
